# The shape of your data

When you plot, you only get so many dimensions. 

You really want to use 2-4 (_maybe 5_) dimensions to keep things. So when you are writing an academic paper you are generally the arbiter of "what is important" by selecting these dimensions.

```{r}
library(ggplot2)
library(palmerpenguins)
data(penguins)
head(penguins)
```

Two dimensions
```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm,
                     color = species)) +
  geom_point()
```

But, if we want to share and explore all di

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm,
                     color = species)) +
  geom_point() +
  facet_grid(~year)
```

Maybe you could squeeze one more in there, but at this point you've kind of lost me

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm,
                     color = species, shape = island)) +
  geom_point() +
  facet_grid(~year)
```


BOOKMARK: IN GGPLOT, EACH DIMENSION IS ITS OWN COLUMN

So when we communicate data we have this tension between sending simple messages with figures while not simplifying/keeping the complexity of the data. 

This is part of allowing a user/viewer to personalize what they see. "Choose your own adventure" type of approach.